---
title: "Troubleshooting Issues"
author: "Johannes Nelson"
date: "2023-12-19"
output: pdf_document
---


# Overview
This document will keep track of some common issues and fixes. I will update it 
as necessary.

## Issues
### 'robotoolbox' package version
As mentioned elsewhere, we are using the development version of the package rather
than the public version, because the public version will not work for our Kobo data.
Eventually, the developer will push his development version to the public. Until 
then, the development version will need to be installed. 

BEFORE executing the Extraction scripts for the first time, run the following 
lines of code:

```{r eval=FALSE}

install.packages("remotes")
remotes::install_gitlab("dickoa/robotoolbox")
```

You should only need to run this one time. After that, the extraction script will
see that a package of that name has been installed and load it normally. If you
already ran the extraction script and installed the public version, you may be 
getting this error:

*Error in rbindlist(drop_nulls(survey), fill = TRUE) :*
*Column 7 of item 3 is length 3 inconsistent with column 1 which is length 5. Only length-1 columns are recycled.*

In this case, you will need to uninstall that package before installing the development
version:

```{r eval=FALSE}
remove.packages('robotoolbox')
install.packages("remotes")
remotes::install_gitlab("dickoa/robotoolbox")
```

You can also try manually deleting the package if the above doesn't work.

### Installing packages
While in most cases, installing packages should run seamlessly without user input,
there might be some cases where it asks for user input while installing packages.

If ever asked: 
```{r eval = FALSE}
"Do you want to install from sources the package which needs compilation? (Yes/no/cancel)"
```
Answer: no


### Relative vs. Absolute filepaths
Calling the scripts with source() should always work with relative filepaths
(filepaths that start with your working directory, beginning with 'Scripts/...').

Be sure that this filepath doesn't start with a '/', as this will tell your 
computer to begin in the system's root directory. Either copy the whole path by
right-clicking on the script and copying the path or simply write the relative
path like this:

```{r eval = FALSE}

source('Scripts/Name_of_script.R')
```
###